---
title: "Introduction to ggforge"
author: "Zaoqu Liu"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Introduction to ggforge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  dpi = 150,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

**ggforge** is a comprehensive visualization framework designed for biomedical and bioinformatics research. This package is greatly inspired by and modified from [plotthis](https://github.com/pwwang/plotthis) by Panwen Wang, with enhancements for high-quality biomedical visualizations.

## Key Features

- 🎨 40+ plotting functions
- 🧬 Specialized for biomedical and bioinformatics research
- 🎯 Unified API design across all functions
- 🌈 Extensive color palette collection
- 🔧 Intelligent type detection and automatic styling
- 📊 Easy multi-panel figure composition

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("Zaoqu-Liu/ggforge")
```

## Loading the Package

```{r library}
library(ggforge)
library(ggplot2)
library(dplyr)
set.seed(8525)
```

# Core Concepts

## Unified API Design

All ggforge plotting functions follow a consistent pattern:

```r
PlotFunction(
  data,                    # Input data frame
  x, y,                    # Main aesthetics
  group_by = NULL,         # Grouping variable
  split_by = NULL,         # Split into multiple plots
  palette = "Paired",      # Color palette
  theme = "theme_ggforge", # Theme
  ...                      # Function-specific parameters
)
```

## Intelligent Type Detection

ggforge automatically detects variable types and applies appropriate styling:

- **Continuous**: Numeric variables → gradient scales
- **Discrete**: Factors/characters → discrete colors
- **Temporal**: Date/DateTime → time axis formatting

## Split and Combine

Create multi-panel figures easily with `split_by`:

```r
LinePlot(..., split_by = "group", combine = TRUE, nrow = 1)
```

# Basic Statistical Plots

## Scatter Plots

Visualize relationships between two continuous variables:

```{r scatter, fig.width=8, fig.height=6}
# Create sample data
scatter_data <- data.frame(
  x = rnorm(100),
  y = rnorm(100),
  group = sample(c("A", "B", "C"), 100, replace = TRUE)
)
scatter_data$y <- scatter_data$y + 0.5 * scatter_data$x

ScatterPlot(
  data = scatter_data,
  x = "x",
  y = "y",
  group_by = "group",
  palette = "npg",
  add_smooth = TRUE,
  add_stat = TRUE,
  title = "Correlation Analysis",
  xlab = "Variable X",
  ylab = "Variable Y"
)
```

**Key parameters:**

- `add_smooth`: Add regression line
- `add_stat`: Display correlation statistics
- `group_by`: Color points by group

## Box Plots

Compare distributions across groups:

```{r boxplot, fig.width=8, fig.height=6}
# Create sample data
box_data <- data.frame(
  group = rep(c("Control", "Treat1", "Treat2"), each = 40),
  value = c(rnorm(40, 10, 2), rnorm(40, 12, 2.5), rnorm(40, 15, 2))
)

BoxPlot(
  data = box_data,
  x = "group",
  y = "value",
  palette = "lancet",
  add_point = TRUE,
  point_alpha = 0.4,
  title = "Treatment Effect Comparison",
  xlab = "Treatment Group",
  ylab = "Expression Level"
)
```

## Violin Plots

Show the full distribution shape:

```{r violinplot, fig.width=8, fig.height=6}
ViolinPlot(
  data = box_data,
  x = "group",
  y = "value",
  palette = "npg",
  add_box = TRUE,
  add_point = TRUE,
  point_size = 0.8,
  point_alpha = 0.4,
  title = "Distribution Comparison"
)
```

## Density Plots

Smooth distribution visualization:

```{r densityplot, fig.width=8, fig.height=6}
DensityPlot(
  data = box_data,
  x = "value",
  group_by = "group",
  palette = "npg",
  add_rug = TRUE,
  title = "Distribution Density",
  xlab = "Expression Value"
)
```

## Bar Plots

Summarize data with error bars:

```{r barplot, fig.width=8, fig.height=6}
BarPlot(
  data = box_data,
  x = "group",
  y = "value",
  palette = "Set2",
  add_errorbar = TRUE,
  errorbar_type = "se",
  title = "Mean Values with Standard Error",
  xlab = "Treatment Group",
  ylab = "Mean Expression"
)
```

## Line Plots

Visualize trends over time or continuous variables:

```{r lineplot, fig.width=9, fig.height=6}
# Create time series data
time_data <- data.frame(
  time = rep(1:10, 3),
  value = c(
    cumsum(rnorm(10, 0.5, 1)),
    cumsum(rnorm(10, 0.3, 1)),
    cumsum(rnorm(10, 0.7, 1))
  ),
  group = rep(c("Group A", "Group B", "Group C"), each = 10)
)

LinePlot(
  data = time_data,
  x = "time",
  y = "value",
  group_by = "group",
  palette = "nejm",
  add_point = TRUE,
  add_smooth = TRUE,
  title = "Time Course Analysis",
  xlab = "Time Point",
  ylab = "Value"
)
```

# Enrichment Analysis

Enrichment analysis is crucial for interpreting high-throughput biological data.

## Enrichment Network

Visualize relationships between enriched terms based on gene overlap:

```{r enrichment-network, fig.width=10, fig.height=8}
# Load example enrichment data
data("enrich_multidb_example")

EnrichNetwork(
  data = enrich_multidb_example,
  top_term = 20,
  layout = "fr",
  palette = "Set3",
  title = "Enrichment Network Analysis"
)
```

**What this shows:**

- Nodes represent enriched terms
- Edges connect terms sharing genes
- Node size indicates significance
- Colors distinguish different databases

## Enrichment Map

Create similarity-based enrichment maps:

```{r enrichment-map, fig.width=10, fig.height=8}
data("enrich_example")

EnrichMap(
  data = enrich_example,
  top_term = 25,
  layout = "fr",
  palette = "Spectral",
  title = "GO Enrichment Map"
)
```

**Interpretation:**

- Clusters of nodes indicate related biological processes
- Distance reflects similarity between terms
- Useful for identifying major functional themes

# GSEA Visualization

Gene Set Enrichment Analysis (GSEA) identifies coordinated changes in pre-defined gene sets.

## GSEA Summary Plot

Overview of multiple GSEA results:

```{r gsea-summary, fig.width=9, fig.height=8}
# Load GSEA example data
data("gsea_example")

GSEASummaryPlot(
  data = gsea_example,
  top_term = 20,
  palette = "RdBu",
  title = "GSEA Results Overview"
)
```

**Understanding the plot:**

- Each row represents a gene set
- Running enrichment score (RES) shown as line plot
- Normalized enrichment score (NES) shown as color
- Quickly identify up/down-regulated pathways

## Individual GSEA Plot

Detailed view of specific pathway enrichment:

```{r gsea-plot, fig.width=10, fig.height=7}
GSEAPlot(
  data = gsea_example,
  gs = gsea_example$ID[1],
  title = gsea_example$Description[1]
)
```

# Single-Cell Analysis

Single-cell RNA-seq reveals cellular heterogeneity at unprecedented resolution.

## Dimensionality Reduction

Visualize cell populations in reduced dimensions:

```{r dimplot, fig.width=9, fig.height=7}
# Load dimension reduction example
data("dim_example")

DimPlot(
  data = dim_example,
  x = "basis_1",
  y = "basis_2",
  group_by = "clusters",
  palette = "igv",
  point_size = 1.5,
  add_label = TRUE,
  label_insitu = TRUE,
  title = "Cell Clustering (UMAP)"
)
```

**Best practices:**

- Use UMAP or t-SNE for visualization
- Color by cluster assignments or cell types
- Add labels for easier interpretation

## Feature Expression

Overlay gene expression on dimensionality reduction:

```{r feature-dimplot, fig.width=9, fig.height=7}
# Create feature data
dim_example$feature <- rnorm(nrow(dim_example))

FeatureDimPlot(
  data = dim_example,
  x = "basis_1",
  y = "basis_2",
  feature = "feature",
  palette = "viridis",
  point_size = 1.5,
  title = "Gene Expression on UMAP"
)
```

## RNA Velocity

Visualize cellular dynamics and differentiation trajectories:

```{r velocity, fig.width=9, fig.height=7}
# Prepare embedding matrices
embedding <- as.matrix(dim_example[, c("basis_1", "basis_2")])
v_embedding <- as.matrix(dim_example[, c("stochasticbasis_1", "stochasticbasis_2")])

VelocityPlot(
  embedding = embedding,
  v_embedding = v_embedding,
  plot_type = "grid",
  title = "RNA Velocity Analysis"
)
```

**Velocity plot types:**

- `raw`: Individual cell velocity vectors
- `grid`: Smoothed velocity field
- `stream`: Streamline visualization

# Genomics Visualization

## Volcano Plot

Standard visualization for differential expression:

```{r volcano, fig.width=9, fig.height=7}
# Create differential expression data
deg_data <- data.frame(
  gene = paste0("Gene", 1:500),
  log2FC = rnorm(500, 0, 1.5),
  pvalue = runif(500, 0, 0.1)
)
deg_data$padj <- p.adjust(deg_data$pvalue, method = "BH")

# Add some significant genes
sig_up <- sample(1:500, 25)
sig_down <- sample(1:500, 25)
deg_data$log2FC[sig_up] <- abs(rnorm(25, 2, 0.5))
deg_data$log2FC[sig_down] <- -abs(rnorm(25, 2, 0.5))
deg_data$padj[c(sig_up, sig_down)] <- runif(50, 0, 0.01)

VolcanoPlot(
  data = deg_data,
  x = "log2FC",
  y = "padj",
  label_by = "gene",
  x_cutoff = 1,
  y_cutoff = 0.05,
  nlabel = 10,
  title = "Differential Expression Analysis",
  xlab = "log2 Fold Change",
  ylab = "-log10(Adjusted P-value)"
)
```

**Customization:**

- Adjust cutoffs for significance
- Control number of labeled genes
- Use custom color schemes

## Manhattan Plot

GWAS and QTL mapping visualization:

```{r manhattan, fig.width=12, fig.height=6}
# Create sample GWAS data
gwas_data <- data.frame(
  chr = rep(paste0("chr", 1:22), each = 500),
  pos = rep(1:500, 22) * 1e5,
  pvalue = runif(11000, 0, 1)
)
gwas_data$pvalue[sample(1:11000, 30)] <- runif(30, 0, 1e-8)

ManhattanPlot(
  data = gwas_data,
  chr_by = "chr",
  pos_by = "pos",
  pval_by = "pvalue",
  threshold = 5e-8,
  title = "Genome-Wide Association Study"
)
```

# Survival Analysis

Kaplan-Meier survival curves are essential for clinical research:

```{r survival, fig.width=9, fig.height=7}
# Create survival data
surv_data <- data.frame(
  time = rexp(150, 0.01),
  status = sample(0:1, 150, replace = TRUE, prob = c(0.4, 0.6)),
  risk = sample(c("Low", "High"), 150, replace = TRUE)
)

KMPlot(
  data = surv_data,
  time = "time",
  status = "status",
  group_by = "risk",
  palette = "jco",
  add_risktable = TRUE,
  add_conf_int = TRUE,
  add_pval = TRUE,
  title = "Overall Survival Analysis",
  xlab = "Time (months)",
  ylab = "Survival Probability"
)
```

**Components:**

- Survival curves with confidence intervals
- Risk table showing numbers at risk
- Log-rank test p-value
- Customizable time points

## Cox Regression Analysis

Multivariate survival analysis with forest plot:

```{r coxplot, fig.width=10, fig.height=8}
# Create sample data with multiple covariates
cox_data <- data.frame(
  time = rexp(200, 0.01),
  event = sample(0:1, 200, replace = TRUE, prob = c(0.3, 0.7)),
  age = rnorm(200, 60, 10),
  gender = sample(c("Male", "Female"), 200, replace = TRUE),
  stage = sample(c("I", "II", "III", "IV"), 200, replace = TRUE),
  treatment = sample(c("A", "B"), 200, replace = TRUE)
)

CoxPlot(
  data = cox_data,
  time = "time",
  event = "event",
  vars = c("age", "gender", "stage", "treatment"),
  plot_type = "forest",
  palette = "nejm",
  title = "Cox Proportional Hazards Model"
)
```

**Key features:**

- Forest plot showing hazard ratios
- Confidence intervals for each covariate
- P-values for statistical significance
- Automatic handling of categorical and continuous variables

# Network Visualization

## Heatmap

Enhanced heatmap for matrix data:

```{r heatmap, fig.width=9, fig.height=8}
# Create sample matrix
set.seed(123)
mat <- matrix(rnorm(100), 10, 10)
rownames(mat) <- paste0("Gene", 1:10)
colnames(mat) <- paste0("Sample", 1:10)

Heatmap(
  data = mat,
  palette = "RdBu",
  title = "Gene Expression Heatmap"
)
```

## Chord Diagram

Visualize relationships between categories:

```{r chord, fig.width=8, fig.height=8}
# Create interaction data
chord_data <- data.frame(
  from = c("CD4 T", "CD8 T", "B cell", "NK", "Monocyte"),
  to = c("Fibroblast", "Endothelial", "Fibroblast", "Tumor", "Tumor"),
  value = c(15, 20, 10, 25, 18)
)

ChordPlot(
  data = chord_data,
  from = "from",
  to = "to",
  y = "value",
  palette = "Set3",
  title = "Cell-Cell Interaction Network"
)
```

## Venn Diagram

Visualize set overlaps:

```{r venn, fig.width=8, fig.height=7}
# Create gene sets
venn_data <- list(
  SetA = paste0("Gene", 1:100),
  SetB = paste0("Gene", 50:150),
  SetC = paste0("Gene", 80:180)
)

VennDiagram(
  data = venn_data,
  palette = "Set2",
  title = "Gene Set Overlap"
)
```

# Color Palettes

ggforge includes extensive color palettes from multiple sources:

```{r palettes, fig.width=12, fig.height=10}
# Show available palettes
show_palettes(head(names(palette_list), 20))
```

**Palette sources:**

- RColorBrewer palettes
- ggsci journal palettes (Nature, Science, NEJM, Lancet, JAMA, JCO)
- viridis color scales
- d3 categorical scales

**Usage tips:**

- Use colorblind-friendly palettes (viridis, cividis)
- Match palette to data type (sequential vs. categorical)
- Consider journal requirements

# Advanced Features

## Multi-Panel Layouts

Create complex multi-panel figures with `split_by`:

```{r split, fig.width=12, fig.height=4}
# Create data with multiple groups
multi_data <- data.frame(
  x = rep(1:10, 9),
  y = rnorm(90, rep(1:3, each = 30), 0.5),
  group = rep(c("A", "B", "C"), each = 30),
  condition = rep(rep(c("Ctrl", "Treat1", "Treat2"), each = 10), 3)
)

LinePlot(
  data = multi_data,
  x = "x",
  y = "y",
  group_by = "condition",
  split_by = "group",
  palette = "nejm",
  add_point = TRUE,
  combine = TRUE,
  nrow = 1
)
```

## Faceting

Use `facet_by` for within-plot faceting:

```{r faceting, fig.width=10, fig.height=6}
# Create faceted data
facet_data <- data.frame(
  x = rep(1:20, 6),
  y = rnorm(120) + rep(rep(c(0, 2, 4), each = 20), 2),
  condition = rep(c("Cond1", "Cond2"), each = 60),
  time = rep(rep(c("Day1", "Day2", "Day3"), each = 20), 2)
)

LinePlot(
  data = facet_data,
  x = "x",
  y = "y",
  group_by = "condition",
  facet_by = "time",
  palette = "jco",
  add_smooth = TRUE
)
```

## Theme Customization

All plots support theme customization:

```{r theme-custom}
# Apply different themes
BoxPlot(
  data = box_data,
  x = "group",
  y = "value",
  palette = "npg",
  theme = "theme_ggforge",
  theme_args = list(base_size = 14)
)
```

# Best Practices

## Data Preparation

**Always ensure your data is in tidy format:**

- One row per observation
- One column per variable
- Each value in its own cell

```{r data-prep, eval=FALSE}
# Good: Tidy format
data_tidy <- data.frame(
  sample = rep(c("S1", "S2", "S3"), each = 3),
  gene = rep(c("GeneA", "GeneB", "GeneC"), 3),
  expression = rnorm(9)
)

# Bad: Wide format
# data_wide <- data.frame(
#   sample = c("S1", "S2", "S3"),
#   GeneA = rnorm(3),
#   GeneB = rnorm(3),
#   GeneC = rnorm(3)
# )
```

## Color Selection

**Guidelines for choosing colors:**

1. **Colorblind-friendly**: Use viridis, cividis, or similar
2. **Consistency**: Use same colors for same categories across figures
3. **Contrast**: Ensure sufficient contrast for readability
4. **Journal requirements**: Check specific requirements

## High-Resolution Export

**Recommended settings for saving figures:**

```{r publication, eval=FALSE}
# High-resolution PDF
ggsave(
  "figure.pdf",
  plot = your_plot,
  width = 8,
  height = 6,
  dpi = 300,
  device = "pdf"
)

# TIFF for some journals
ggsave(
  "figure.tiff",
  plot = your_plot,
  width = 8,
  height = 6,
  dpi = 600,
  device = "tiff",
  compression = "lzw"
)
```

## Reproducibility

**Always set seed for reproducible results:**

```{r reproducibility, eval=FALSE}
set.seed(8525) # ggforge default seed
```

# Troubleshooting

## Common Issues

### Issue 1: Column not found

```{r troubleshoot1, eval=FALSE}
# Check column names (case-sensitive)
colnames(your_data)
```

### Issue 2: Palette not working

```{r troubleshoot2, eval=FALSE}
# See available palettes
names(palette_list)

# Or use custom colors
custom_colors <- c("#1F77B4", "#FF7F0E", "#2CA02C")
```

### Issue 3: Legend position

```{r troubleshoot3, eval=FALSE}
# Control legend
YourPlot(..., theme_args = list(legend.position = "bottom"))
```

# Getting Help

- **Documentation**: `?FunctionName` or `help(FunctionName)`
- **Examples**: `example(FunctionName)`
- **GitHub Issues**: https://github.com/Zaoqu-Liu/ggforge/issues
- **Email**: liuzaoqu@163.com

# Acknowledgments

This package is greatly inspired by and modified from [plotthis](https://github.com/pwwang/plotthis) by Panwen Wang. We are deeply grateful for the original work and design philosophy that made ggforge possible.

We also acknowledge:
- [ggplot2](https://ggplot2.tidyverse.org/) for the grammar of graphics
- [patchwork](https://patchwork.data-imaginist.com/) for multi-panel composition
- The entire R community for their excellent packages

# Session Info

```{r sessioninfo}
sessionInfo()
```
